Changes in accessibility to food banks and food services during COVID-19 and implications for low income populations in Hamilton, Ontario

### 2021-04-24 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.Rmd
Push my knit

```{r install-data-package, include = FALSE}
# Run only once if needed to install data package `sobiEquity`
if (!require("foodbanksHamilton", character.only = TRUE)) {
  devtools::install_github("https://github.com/paezha/Accessibility-Food-Banks-Hamilton", subdir = "foodbanksHamilton")
}
```

>>>>

```{r install-data-package, include = FALSE}
# Run only once if needed to install data package 
if (!require("foodbanksHamilton", character.only = TRUE)) {
  devtools::install_github("https://github.com/paezha/Accessibility-Food-Banks-Hamilton", subdir = "foodbanksHamilton")
}
```

### 2021-04-24 Accessibility-Foodbanks-Hamilton/Accessibility-Foodbanks-Hamilton.tex
Push my knit

The locations of food banks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. While some food bank
services have a specific a target population, such as prioritizing
family with young children aged between 0 and 3 or accepting only those
providing proof of low-income status through housing and utility
statements, all the food bank services indicated below are designed to
accommodate those in need of food at zero to low cost. With our focus on
food banks and food services that offer free or low-cost meals at
particular locations, we first removed services such as Meals on Wheels
and other food access services such as food box, community kitchens,
student nutrition programs, and shopping and transportation. In
addition, two free meal services held on different days at the same
location were collapsed into a single service point for the
accessibility analysis. Additional details on the operations of
individual facilities is not publicly available, and with the changes in
operations it proved unfeasible to collect it. For this reason, the
analysis to follow is of accessibility to the location of food banks and
services, but not to specific services (e.g., breakfasts vs.~food
boxes).

@@

\noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the food bank. In other words, the expected travel time
is the weighted sum of travel times to the food bank, with the weights
given by the expected modal split in the TAZ.

@@

For the more suburban clusters of zones, the decrease in accessibility
is derived from the closure of locations throughout the city reachable
by car. In the cluster of central suburban zones for example, low-income
households in the outer ring of zones that exhibit medium to high
decreases in accessibility within this cluster appear to be largely
auto-dependent in their tripmaking, which each each exhibiting between
85-100\% of their modal split for car trips. This results in the parcels
within these zones having a large number of potentially accessible
locations in the travel time matrix. But by extension, the change in
accessibility over the pre- and during-COVID-19 time periods is affected
not only by the closure of service locations proximate to the zones, but
also the locations in the central city. The zone with the greatest
decrease in accessibility within this cluster
(\ensuremath{-9\times 10^{-4}}) has a high rate of car trips and
connects to the most food bank locations in total as well as those that
stayed open or closed. The exception to this is the zone in the middle
of this cluster, which has a much higher proportion of transit use 32\%.
This modal split weighting results in fewer food bank and service
locations accessible from the parcels within this zone which, by
extension, limits their access to closed locations reachable within 15
minutes in our calculations.

@@

Just as the effects of the closures appear to have been uneven in space,
they also seem to have differently impacted various population segments.
Table \ref{tab:accessibility-by-age} shows the estimated number of
people in each age group by their level of accessibility before and
during the pandemic. Here, it is important to note that the quartiles
are relative: people in the top 25\% of accessibility still have lower
accessibility during the pandemic than before. In reality, every
population group is worse off during the pandemic in terms of their
accessibility to food banks and services in the City of Hamilton.
However, some age groups were affected more. As seen in the table, the
distribution within quartiles for adults changed only slightly, and in
fact there are fewer adults in the bottom three quartiles as some found
themselves in the (deteriorated) top quartile of accessibility during
the pandemic. The story is different for seniors, a greater number of
whom are now in the third quartile, due to a combination of seniors
facing a worse accessibility situation, and just a few moving from the
bottom quartile (pre-pandemic) to the third during the pandemic. Those
aged 18 and less appear to have fared worse, and the largest change is
in the number of children who were in the third quartile before the
pandemic found themselves in the bottom accessibility class during the
pandemic. These results suggest that, through a combination of the
typical modes of transportation of lower income households, and the
spatial distribution of the population, the food bank/service locations
closed had a differential impact that more greatly affected the youngest
among the population in low income households.

@@

@@

>>>>

The locations of food banks and related food services were obtained from
the Hamilton Public Library's Food Access Guide\footnote{http://foodaccessguide.ca/sites/default/files/partnersites/pdf/foodaccessguide.pdf}.
The guide was updated in April of 2021 to indicate any change affected
on the services due to the pandemic. This includes modified business
hours, a need to make reservations before frequenting, and locations
that have completely shut down in consequence. Table 1 defines each
service type and the number of locations pre- and during the COVID-19
pandemic. While some food bank services have a specific a target
population, such as prioritizing families with young children aged
between 0 and 3 or accepting only those providing proof of low-income
status through housing and utility statements, all the food bank
services indicated below are designed to accommodate those in need of
food at zero to low cost. With our focus on food banks and food services
that offer free or low-cost meals at particular locations, we first
removed services such as Meals on Wheels and other food access services
such as food box, community kitchens, student nutrition programs, and
shopping and transportation. In addition, two free meal services held on
different days at the same location were collapsed into a single service
point for the accessibility analysis. Additional details on the
operations of individual facilities is not publicly available, and with
the changes in operations it proved unfeasible to collect it. For this
reason, the analysis to follow is of accessibility to the location of
food banks and services, but not to specific services (e.g., breakfasts
vs.~food boxes).

@@

\noindent where \(p^k_i\) is the proportion of trips by mode \(k\) in
the TAZ of parcel \(i\), and \(tt^k_{ij}\) is the travel time from
parcel \(i\) to the food bank. In other words, the expected travel time
is the weighted sum of travel times to the food bank, with the weights
given by the expected modal split of trips made by low-income households
in the TAZ per the TTS data.

@@

For the more suburban clusters of zones, the decrease in accessibility
is derived from the closure of locations throughout the city reachable
by car. In the cluster of central suburban zones for example, low-income
households in the outer ring of zones that exhibit medium to high
decreases in accessibility within this cluster appear to be largely
auto-dependent in their tripmaking, which each each exhibiting between
85-100\% of their modal split for car trips. This results in the parcels
within these zones having a large number of potentially accessible
locations in the travel time matrix. But by extension, the change in
accessibility over the pre- and during-COVID-19 time periods is affected
not only by the closure of service locations proximate to the zones, but
also the locations in the central city. The zone with the greatest
decrease in accessibility within this cluster (-0.0009) has a high rate
of car trips and connects to the most food bank locations in total as
well as those that stayed open or closed. The exception to this is the
zone in the middle of this cluster, which has a much higher proportion
of transit use 32\%. This modal split weighting results in fewer food
bank and service locations accessible from the parcels within this zone
which, by extension, limits their access to closed locations reachable
within 15 minutes in our calculations.

@@

Just as the effects of the closures appear to have been uneven in space,
they also seem to have differently impacted various population segments.
Table \ref{tab:accessibility-by-age} shows the estimated number of
people in each age group by their level of accessibility before and
during the pandemic. Here, it is important to note that the quartiles
are relative: people in the top 25\% of accessibility still have lower
accessibility during the pandemic than before. In reality, every
population group is worse off during the pandemic in terms of their
accessibility to food banks and services in the City of Hamilton.
However, some age groups were affected more. As seen in the table, the
distribution within quartiles for adults changed only slightly, and in
fact there are fewer adults in the bottom three quartiles as some found
themselves in the (deteriorated) top quartile of accessibility during
the pandemic. The story is different for seniors, a greater number of
whom are now in the third quartile, due to a combination of seniors
facing a worse accessibility situation, and just a few moving from the
bottom quartile (pre-pandemic) to the third during the pandemic. Those
aged 18 and less appear to have fared worse, and the largest change is
in the number of children who were in the third quartile before the
pandemic and found themselves in the bottom accessibility class during
the pandemic. These results suggest that, through a combination of the
typical modes of transportation of lower income households and the
spatial distribution of the population, the food bank/service locations
closed had a differential impact that more greatly affected the youngest
among the population in low income households.

@@

We use data from the census to model low-income households as a proxy
for households that are possibly experiencing food insecurity. However,
the degree to which our low-income cut-off of CAD\$40,000 reflects food
insecurity in the different zones of the study area is not known.

While the travel survey allows us to model multi-modal accessibilities
that align with travel behaviour observed in the travel survey and
capture differences in accessibility by age categories, the use of
travel survey data for modelling food insecurity also has its
limitations. Research into the population weighting methods used in the
TTS note that the survey may under-count the lowest- and highest-income
households in the survey study region, although the magnitude of this
under-counting is unknown as approximately 20\% of respondents to the
2016 survey did not report their income (Rose, 2018). In that regard,
the modal splits of low-income households observed through the TTS data
may not accurately reflect the travel behaviours of food insecure
populations.

In the absence of information on how food insecure households travel to
food banks and related services, we adopted a normative approach that
examines accessibility to food banks reachable within 15 minutes based
on standards for accessible drinking water\ldots{}

@@

\leavevmode\hypertarget{ref-rose2018transportation}{}%
Rose, A., 2018. Transportation tomorrow survey 2016: Data expansion and
validation. Data Management Group, Department of Civil Engineering,
University of Toronty, Malatest \& Associates.
